Structuring Theories with Implicit Morphisms

### 2018-10-19 mmt.tex
colons
finished for submission

\paragraph{Flat Modules}
\mmt provides formal syntax for both structured modules and module expressions and defines their semantics via \textbf{flattening}, which defines for every module $M$ a flat module $\flt{M}$.

\textbf{Flat theories} are lists of declarations $c:E[=e]$ where $E$ and $e$ are expression, and the latter is optional.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$ (see below for the definition of well-formed expressions).
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=P$, which establish theorem $F$ via proof $P$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

Individual formal languages arise as fragments of \mmt: they single out the well-formed expressions by defining the two \mmt-\textbf{judgments} $\der_T e:e'$ (typing) and $\der_T e\equiv e'$ (equality) for every theory $T$ and $e,e'\in\Exp{T}$.
%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
The details can be found in \cite{rabe:howto:14}.

@@

An \mmt \textbf{diagram} consists of a set of named structured module declarations.
\begin{nomodexp}
For a given diagram, we write $\Thy$ for the set of theory names.
And we write $\Mo{S}{T}$ for the set of morphisms defined by
\begin{compactitem}
 \item for every declaration $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$,
 \item for every $T\in\Thy$, we have $\id{T}\in\Mo{T}{T}$,
 \item for every $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we have $M;N\in\Mo{R}{T}$.
\end{compactitem}
$\Thy$ and $\Mo{S}{T}$ form the category of theories and morphisms.
\end{nomodexp}
\begin{modexp}
Every diagram induces sets of anonymous module expressions.
These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
Every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
\end{modexp}
Then flattening assigns
\begin{compactitem}
 \item to each $T\in\Thy$ the flat theory $\flt{T}$,
 \item to each $M\in\Mo{S}{T}$ the flat morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
\end{compactitem}

@@

\mmt itself is independent of the base logic and provides a theory and theory morphism layer on top of an arbitrary declarative language.
Our results in this paper are similarly logic-independent.
The syntax and semantics of that base logic is provided by a set of rules for the judgments $\vdash_T e:E$ and $\vdash_T e=e'$ for $e,e',E\in\Exp{T}$.
The declarations in theories and morphisms are subject to typing conditions using these judgments.
The main theorem about \mmt is that well-typed morphisms $M:S\to T$ preserve all judgments, e.g,., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
This includes the preservation of truth via the propositions-as-types principle if $E$ is a proposition and $e$ its proof.

The only aspect of the inference system for these judgments that is relevant to our purposes here is the typing rule for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
Here, in order to avoid case distinctions for the case where a definition is present or not, we use a combined typing+equality judgment: $e_1=e_2:E$ represents the conjunction of $e_1:E$, $e_2:E$, and $e_1=e_2$.
Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
\[M(c) = \cas{e\mifc (c:E) \in \flt{S}, (c:=e)\in\flt{M} \\ M(e) \mifc (c:E=e)\in \flt{S}}\]
Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
\footnote{The \mmt tool accepts $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition. In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$. This is important for efficiency but not essential for our purposes here.}

Note that these base cases introduce a mutual recursion between well-formedness and flattening:
the well-formedness of a declaration in a theory depends on the flattening of all preceding declarations;
correspondingly, the well-formedness of an assignment in a morphism depends on the homomorphic extension of the morphism obtained by flattening of all preceding assignments.
Vice versa, well-formedness is a precondition for defining the flattening --- the definition of flattening may become nonsensical if applied to ill-formed modules.

It is desirable to define well-formedness independently of flattening.
But the mutual recursion makes sense from an implementation perspective: typical tools for formal systems first parse, check, and flatten a declaration entirely before moving on to the next declaration.
Moreover, often the checking of well-formedness is as difficult or expensive as flattening anyway.
In fact, inspecting practical systems with modular features (which often do not have a full rigorous formal specification) shows that our definition elegantly captures the essential commonalities between them.

@@

\begin{definition}[Theory]\label{def:theory}
The grammar for theories and expressions is of the form
\begin{grammar}
TDec     & T=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
Dec      & n:E\opt{=E}           & \text{constant declaration}\\
c        & t?n                   & \text{qualified constant identifiers} \\
E        & c \alt \ldots         & \text{expressions built from constants} \\
\end{grammar}

In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions over the previously introduced constants only.
We omit the remaining productions for expressions here, which allow forming complex expressions using application, binding, variables, literals, etc.
\end{definition}

\begin{example}\label{ex:thgroup}
For the purposes of our running examples, we assume a fixed base logic that provides a first-order logic a simple type system.
\lstinline|type| is the universe of types, \lstinline|A ⟶ B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
We also use \mmt notations, which are attached to constant declarations as \lstinline|# <not>|; these are omitted from the formal grammar above because we only use them in the examples.

Then the (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
\begin{mmtcode}
Group =
  U		: type
  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit		: U
  inverse 	: U ⟶ U   # 1 ⁻¹ 
  // axioms omitted
\end{mmtcode}
\begin{mmtcode}
DivisionGroup =
  U 	: type
  div 	: U ⟶ U ⟶ U  # 1 / 2 
  unit 	: U
  // axioms omitted
\end{mmtcode}
\end{example}

At this point, this grammar only allows forming flat theories, i.e., $\Thy=\thy$.
It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding productions for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
This makes it easy to incrementally add new declaration kinds and operators to the language.

%These theories are similar to OpenMath content dictionaries \cite{openmath}.
%In particular, \mmt uses OpenMath-style qualified symbol identifiers $t?n$.%
%\footnote{Of course, the \mmt IDE and web interface parse and display these identifiers only as $n$ (if that is unambiguous), especially if they are used in formulas.
%But the interface always knows the qualified identifier and shows it, e.g., on hover.}
%Thus, symbols of the same name declared in different atomic theories are always distinguished.

We proceed accordingly for flat morphisms:

\begin{definition}[Morphism]\label{def:morphism}
A \textbf{morphism} is of the form
\begin{grammar}
MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{flat morphism declaration}\\
Ass    & c:=o                            & \text{assignment to symbol}\\
M      & m \alt \id{T} \alt M;M          & \text{morphism expressions}
\end{grammar}
such that a morphism declaration contains exactly one assignment $c:=o$ for each $c\in\dom{S}$, all of which satisfying $o\in\Exp{T}$.
\end{definition}

\begin{example}[Morphisms]\label{ex:dg2g}
We give the morphism \cn{DG2G} between the theories from Ex.~\ref{syn:incl}:
\begin{mmtcode}
DG2G : DivisionGroup -> Group =
  U := U
  div := [a,b] a ∘ (b⁻¹) 
  unit := unit
  // assignments to axioms omitted
\end{mmtcode}
It maps the universe and unit of a division group to the corresponding notions of a group.
And we have $\cn{DG2G}(a/b)=a\circ b^{-1}$.
Additionally, the morphism must maps every axiom of \cn{DivGroup} to a proof in \cn{Group} of the translated statement, but we omit those assignments.
\end{example}

@@

\begin{definition}[Diagram]
A \textbf{diagram} is a list of theory and morphism declarations:
\begin{grammar}
Dia    & \rep{(TDec \alt MDec)}          & \text{diagrams}\\
\end{grammar}
A diagram is well-formed if each atomic theory/morphism declaration has a unique name and is well-formed relative to the diagram preceding it.
\end{definition}

We give some examples how to add modularity principles to the grammar:

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
This allows writing the theory $\cn{Group}$ from Ex.~\ref{ex:thgroup} by extending a theory of monoids.
Again omitting all axioms, this looks as follows:
\begin{mmtcode}
Monoid =
  U: type
  op : U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit	: U
Group =
  include Monoid
  inverse : U ⟶ U   # 1 ⁻¹ 
\end{mmtcode}
\end{example}

@@

For each example, we will define the semantics in the next section.

@@

\paragraph{Flattening}
Just like the syntax, the definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added later independently of each other.
In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $\flt{Y}$.

For theories $T$, at this point, this induction only has the base case of a reference to a declared theory $t$.
Further cases arise when we add structuring declarations.

For the base case, of a theory $t=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
\[\flt{t}=\flt{\Sigma}\]
where
 \[\flt{\cdot}=\es\]
 \[\flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
Here $\fltd{T}{D}$ is the flattening of declaration $d$ relative to $\flt{\Sigma}$.

At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{(n:E[=e])}=\{t?n:E[=e]\}\]
where $t$ is the name of the containing theory.

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{(c:=e)}=\{c:=e\}\]
\begin{nomodexp}
Moreover, for $T\in\Thy$ we define
\[\flt{\id{T}}=\{c:=c \;|\; c:E\in\flt{T}\}\]
And for $M;N\in\Mo{R}{T}$, we define
\[\flt{(M;N)}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c:E\in\flt{R}\}\]
Note that in both cases, we only have to consider $R$-constants without definiens.
\end{nomodexp}

@@

Note that the definition of $\fltd{\Sigma}{n:E=e}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
Therefore, includes can never lead to name clashes.

Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
Therefore, flattening and well-formedness must be distinguished:
  \[\fltd{\sigma}{(\icl{M})} = \flt{M}\]
under the condition $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.

The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from Ex.~\ref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers from $\cn{Monoid}$ -- i.e. the prefixes $t$ in $t?n$.
\end{example}

>>>>

\paragraph{Flat Modules}
\textbf{Flat theories} are lists of declarations $c:E[=e]$ where $E$ and $e$ are expressions, and the latter is optional.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$.
Every theory induces the set $\Exp{T}$ for the set of closed \textbf{expressions} using only the symbols $c\in\dom{T}$.
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=P$, which establish theorem $F$ via proof $P$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

@@

A \textbf{diagram} consists of a set of theory and morphism declarations.
\begin{nomodexp}
For a given diagram, we write $\Thy$ for the set of declared \textbf{theories}.
And we write $\Mo{S}{T}$ for the set of \textbf{morphisms} defined by
\begin{compactitem}
 \item for every declaration $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$,
 \item for every $T\in\Thy$, we have $\id{T}\in\Mo{T}{T}$,
 \item for every $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we have $M;N\in\Mo{R}{T}$.
\end{compactitem}
$\Thy$ and $\Mo{S}{T}$ form the category of theories and morphisms.
\end{nomodexp}
\begin{modexp}
Every diagram induces sets of anonymous module expressions.
These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
Every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
\end{modexp}

The theories and morphisms in a diagram may be structured (e.g., by using \emph{include} declarations), and \mmt defines their semantics via \textbf{flattening}, which defines for each $T\in\Thy$ the flat theory $\flt{T}$, and for each $M\in\Mo{S}{T}$ the flat morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.

@@

\mmt is independent of the base logic and provides a theory and theory morphism layer on top of an arbitrary declarative language.
Individual logics arise as fragments of \mmt: they single out the well-formed expressions by defining the \textbf{judgment} $\der_T e=e':E$ for typing and equality.
(We treat the plain typing judgment $\der_T e:E$ as the special case $\der_T e=e:E$.)
%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
The declarations in theories and assignments in morphisms are subject to typing conditions, and the main theorem about \mmt is that well-typed morphisms $M:S\to T$ preserve all judgments, i.e., if $\vdash_S e=e':E$, then $\vdash_T M(e)=M(e'):M(E)$.
This includes the preservation of truth via the propositions-as-types principle if $E$ is a proposition and $e$ its proof.

We do not give all rules for these judgments.
The only rule that is relevant to our purposes here is the one for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
(Here, we merge the two cases where $c$ has a/no definiens $e$ into one rule for convenience.)
Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
\[M(c) = \cas{e\mifc (c:E) \in \flt{S}, (c:=e)\in\flt{M} \\ M(e) \mifc (c:E=e)\in \flt{S}}\]
Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
\footnote{The \mmt tool accepts $(c:=e)\in \flt{M}$ even if $(c:E=e')\in\flt{S}$. In that case, \mmt checks $\vdash_T M(e')=e:M(E)$ and puts $M(c)=e$. This is important for efficiency but not essential for our purposes here.}

These base cases introduce a mutual recursion between well-formedness and flattening:
the well-formedness of a declaration in a theory depends on the flattening of all preceding declarations;
correspondingly, the well-formedness of an assignment in a morphism depends on the homomorphic extension of the morphism obtained by flattening of all preceding assignments.
Vice versa, well-formedness is a precondition for defining the flattening --- the definition of flattening may become nonsensical if applied to ill-formed modules.
It might be desirable to define well-formedness independently of flattening.
But our definition captures the typical behavior of practical systems, which first parse, check, and flatten one declaration entirely before moving on to the next one.

@@

\begin{definition}[Theory]\label{def:theory}
The grammar for theories and expressions is
\begin{grammar}
TDec     & T=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
Dec      & n:E\opt{=E}           & \text{constant declaration}\\
c        & T?n                   & \text{qualified constant identifiers} \\
E        & c \alt \ldots         & \text{expressions built from constants} \\
\end{grammar}

In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions over the previously introduced constants only.
We omit the productions and typing rules for the remaining expressions, which include application, binding, variables, literals, etc.
\end{definition}

\begin{example}\label{ex:thgroup}
For the purposes of our running examples, we assume a fixed base logic that provides a simple type system.
\lstinline|type| is the universe of types, \lstinline|A ⟶ B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
We also use \mmt notations, which are attached to constant declarations as \lstinline|# <not>|; these are omitted from the formal grammar above because we only use them in the examples.

Then the (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
\begin{mmtcode}
Group =
  U       : type
  op      : U $\to$ U $\to$ U	 # 1 $\circ$ 2 
  unit    : U
  inverse : U $\to$ U   # 1 $^{-1}$ 
  // axioms omitted
\end{mmtcode}
\begin{mmtcode}
DivGroup =
  U   : type
  div : U $\to$ U $\to$ U  # 1 / 2 
  unit: U
  // axioms omitted
\end{mmtcode}
\end{example}

%
%These theories are similar to OpenMath content dictionaries \cite{openmath}.
%In particular, \mmt uses OpenMath-style qualified symbol identifiers $t?n$.%
%\footnote{Of course, the \mmt IDE and web interface parse and display these identifiers only as $n$ (if that is unambiguous), especially if they are used in formulas.
%But the interface always knows the qualified identifier and shows it, e.g., on hover.}
%Thus, symbols of the same name declared in different atomic theories are always distinguished.
%
We proceed accordingly for flat morphisms:

\begin{definition}[Morphism]\label{def:morphism}
The grammar for \textbf{morphisms} is
\begin{grammar}
MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{flat morphism declaration}\\
Ass    & c:=E                            & \text{assignment to symbol}\\
M      & m \alt \id{T} \alt M;M          & \text{morphism expressions}
\end{grammar}
A morphism must contain exactly one assignment $c:=e$ for each $(c:E)\in\flt{S}$, all of which must satisfy $\der_T e: M(E)$.
\end{definition}

\begin{example}[Morphisms]\label{ex:dg2g}
We give the morphism \cn{DG2G} between the theories from Ex.~\ref{syn:incl}:
\begin{mmtcode}
DG2G : DivGroup -> Group =
  U     := U
  div 	:= [a,b] a $\circ$ (b$^{-1}$) 
  unit 	:= unit
  // assignments to axioms omitted
\end{mmtcode}
It maps the universe and unit of a division group to the corresponding notions of a group.
And we have $\cn{DG2G}(a/b)=a\circ b^{-1}$.
Additionally, the morphism maps every axiom of \cn{DivGroup} to a proof in \cn{Group} of the translated statement, but we omit those assignments.
\end{example}

@@

\begin{definition}[Diagram]
A \textbf{diagram} is a list of theory and morphism declarations:
\begin{grammar}
Dia    & \rep{(TDec \alt MDec)}          & \text{diagrams}\\
\end{grammar}
Each theory/morphism declaration must have a unique name and be well-formed relative to the diagram preceding it.
\end{definition}

At this point, the grammar only allows forming flat theories and morphisms.
\begin{modexp}
It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding productions for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
\end{modexp}
Structuring features can be added to the language incrementally and individually.
\begin{modexp}
We give some examples how to add modularity principles to the grammar:
\end{modexp}

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{T} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
This allows writing the theory $\cn{Group}$ from Ex.~\ref{ex:thgroup} by extending a theory of monoids.
Again omitting all axioms, this looks as follows:
\begin{mmtcode}
Monoid =
  U 	: type
  op 	: U $\to$ U $\to$ U	 # 1 $\circ$ 2 
  unit	: U
Group =
  include Monoid
  inverse : U $\to$ U   # 1$^{-1}$ 
\end{mmtcode}
\end{example}

@@

@@

\paragraph{Flattening}
The definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added independently of each other.
\begin{modexp}
In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $\flt{Y}$. 
For theories $T$, at this point, this induction only has the base case of a reference to a declared theory $t$.
Further cases arise when we add structuring declarations.
\end{modexp}

\begin{definition}[Flattening]
For the base case, of a theory $t=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
\[\flt{t}=\flt{\Sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb
 \flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
Here $\fltd{T}{D}$ is the flattening of declaration $D$ relative to $\flt{\Sigma}$.
At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{(n:E[=e])}=\{t?n:E[=e]\}\]
where $t$ is the name of the containing theory.

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, $\flt{m}$ is defined by induction on the assignments in $\sigma$:
\[\flt{m}=\flt{\sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{(c:=e)}=\{c:=e\}\]
\begin{nomodexp}
Moreover, for $T\in\Thy$ we define $\flt{\id{T}}=\{c:=c \;|\; (c:E)\in\flt{T}\}$
And for $M;N\in\Mo{R}{T}$, we define $\flt{(M;N)}=\{c:=\flt{N}(\flt{M}(c)) \;|\; (c:E)\in\flt{R}\}$.
Note that in both cases, we only have to consider constants without definiens.
\end{nomodexp}
\end{definition}

@@

Note that the definition of $\fltd{\Sigma}{n:E=e}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
Therefore, includes can never lead to name clashes.
Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.

The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
Therefore, flattening is not always defined:
  \[\fltd{\sigma}{(\icl{M})} = \flt{M}\]
and $\sigma,\icl{M}$ is well-formed $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
\end{example}

\begin{example}[Continuing Ex.~\ref{syn:incl}]
$\cn{Group}^\flat$ is obtained by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in the theory as given in Ex.~\ref{ex:thgroup} except that the identifiers are now, e.g., $\cn{Monoid}?U$ instead of $\cn{Group}?U$.
\end{example}
